---
title: "Data Visualisation Workshop-1"
output:
  html_document:
    df_print: paged
---
Load the R-packages required for completing this workshop. If you do not have any of the R-packages installed on your machine, please use `install.packages("pkgname")` to install the R-package `pkgname` from CRAN. To complete this workshop, fill in all `...` with correct R object names.

```{r, eval=FALSE}
# Load required R libraries
library(dslabs) # for murders dataset
library(tidyverse) # for ggplot2 and other data wrangling packages
library(waffle) # to create waffle chart
library(RColorBrewer) # for selecting attractive color palette
library(ggthemes) # for attractive visual themes
library(ggridges) # for ridge density plot
library(viridis) # for some interesting colour scheme
```

Load the `murders` data and assign it to a variable called murders.

```{r, eval=FALSE}
# Load the murders data
data("murders")
# Assign murders data to a variable called murders
murders <- murders
# Get a glimpse of the data
dplyr::glimpse(murders)
# Get the summary of the data
summary(murders)
```

### Q1. Create a **Pie chart** showing the regional murder percentages

First create a `data.frame` with region-specific murder percentages. Check that the percentages for the four regions add to 100.

```{r, eval=FALSE}
# group by region and compute the total number of cases for each region
murders_regional_summary <- murders %>%
                      dplyr::group_by(region) %>%
                      summarise(total = sum(total), .groups = "drop") %>%
                      mutate(percentage = round((total/(sum(total))) * 100))

# print murders_regional_summary
murders_regional_summary

# Check that sum of all percentages is 100
sum(murders_regional_summary$percentage)
```

Create your first pie chart. Note that we are using `geom_col()` in the reference to the polar coordinates to create the pie chart. The function `geom_col()` is used for creating stacked-bar plots, so the pie-chart can be thought of a transformed stacked-bar chart, transformed using the `coord_polar()` function.

```{r, eval=FALSE}
# Create the barebone Pie chart using murders_regional_summary
murders_regional_summary %>% 
  ggplot(aes(x="", y=percentage, fill=region)) +
  geom_col() +
  coord_polar("y", start=0, direction=-1)
```

Let's polish our Pie chart little bit and create a more presentable visualisation.

Let's change the theme first.

```{r, eval=FALSE}
# Add theme_void()
murders_regional_summary %>% 
  ggplot(aes(x="", y=percentage, fill=region)) +
  geom_col() +
  coord_polar("y", start=0, direction=-1) +
  theme_void()
```

Now we have a much cleaner visualisation, but we're missing the key data in our visualisation. We are not sure what percentage each pie represents.

Let's add some annotations inside the color wheel to show the corresponding percentage values.

```{r, eval=FALSE}
# Add geom_text() --- arguments are given
murders_regional_summary %>% 
  ggplot(aes(x="", y=percentage, fill=region)) +
  geom_col() +
  coord_polar("y", start=0, direction=-1) +
  theme_void() +
  geom_text(aes(label = paste0(percentage, "%")),
            position = position_stack(vjust = 0.5), size=10) 
  
```

Let's put the finishing touches.

```{r, eval=FALSE}
# 1. Order the regions in the decreasing order of percentages
# Change the fill mapping from region to fct_reorder(region, -percentage)
# 2. Create a proper legend title: "Region"
# Use labs(fill = "Region")
# 3. Choose an appropriate plot title
# ggtitle("Regional breakdown of 2010 gun murders")
# 4. Choose a better color palette for Regions
# Use scale_fill_brewer(palette = "Accent")
# 5. The theme() elements will discussed in the next two lectures 
murders_regional_summary %>% 
  ggplot(aes(x="", y=percentage, fill=fct_reorder(region, -percentage))) +
  geom_col() +
  coord_polar("y", start=0, direction=-1) +
  theme_void() +
  geom_text(aes(label = paste0(percentage, "%")),
            position = position_stack(vjust = 0.5), size=10) +
  labs(fill = "Region") +
  ggtitle("Regional breakdown of 2010 gun murders") +
  scale_fill_brewer(palette = "Accent") +
      theme(legend.title = element_text(size=18, hjust = 0.5),
        legend.text = element_text(size=15),
        legend.key.size = unit(1, 'cm'),
        plot.title = element_text(hjust = 0.5, size=15, face = "bold"))
```


### Q2. Create a **Waffle chart** showing the regional murder percentages

```{r, eval=FALSE}
# Sort percentage values in the murders_regional_summary 
murders_regional_summary2 <- murders_regional_summary %>% 
                             arrange(desc(...))

# We need the vector of percentage values to create the waffle chart
# Assign the vector to percent_by_region
percent_by_region <- murders_regional_summary2$...

# percent_by_region should be a named vector --- assign the regions to names 
names(percent_by_region) <- murders_regional_summary2$...

# Create the waffle chart using waffle() function
# Assign percent_by_region to the first argument
# Assign number of rows
# Assign brewer.pal(n=4, name="Accent") to colors

waffle::waffle(..., rows= ..., colors = ...) +
  theme(legend.title = element_text(size=15, hjust = 0.5),
        legend.text = element_text(size=12),
        plot.title = element_text(hjust = 0.5, size=15,
                                  face = "bold"))  +
  ggtitle("Waffle chart: Regional breakdown of 2010 gun murders")
```

### Q3. Create a **Donut chart** showing the regional murder percentages

```{r, eval=FALSE}
# Add xlim(0.5, 2.5) to the code of Pie Chart
# Use murders_regional_summary as the data
# Put x = 2 and y = percentage into mappig aesthetics
# Choose appropriate fill aesthetics
# assign color="black"
# Choose appropriate values for the arguments of the coord_polar() function
# Choose an appropriate theme
# Assign appropriate legend title fill = "Region"
# Assign palette = "Accent" in scale_fill_brewer()

... %>% ggplot(aes(x=..., y=..., fill=fct_reorder(..., ...))) +
  geom_col(color=...) +
  coord_polar(..., start=..., direction=...)+
  geom_text(aes(label = paste0(percentage, "%")),
            position = position_stack(vjust = 0.5),
            size=8) +
  theme_...() +
  labs(fill = ...) +
  theme(legend.title = element_text(size=15, hjust = 0.5),
        legend.text = element_text(size=12),
        legend.key.size = unit(1, 'cm'),
        plot.title = element_text(hjust = 0.5, size=15,
                                  face = "bold")) +
  scale_fill_brewer(... = ...) +
  ggtitle("Donut chart: Regional breakdown of 2010 gun murders") +
  ...
```

Let's now create a stacked bar chart using the WHO's disease dataset. First we need to read the data into our R session.

```{r, eval = FALSE}
# Read in the csv dataset: world_health_org_disease_data.csv
disease_data <- read.csv("world_health_org_disease_data.csv"))

# Summary of the WHO Disease data
summary(...)
```

We can see that some of the variables in the `disease_data` are stored as
`character` vectors --- we should be transform these variables into `factor` variables.

```{r, eval=FALSE}
# Use is.character and as.factor function in mutate_if() to
# transform all character vectors into factor variables.
disease_data <- disease_data %>%
                mutate_if(..., ...)

# Take a glimpse to check if the transformations happened
glimpse(...)

# Summary of the data
summary(...)

# Create disease_data2016 by selecting data corresponding to year = 2016
... <- disease_data %>%
                    filter(...)

# Summary of the disease_data2016
summary(...)
```


We have six regions. Let's look at the distribution of these diseases in the
six regions and compare them. 

```{r, eval = FALSE}
# Group by region and disease type and compute the total number of cases
disease2016_summary <- disease_data2016 %>%
                       group_by(...) %>%
                       summarise(cases = ..., 
                                 .groups = "drop")
```

### Q4. Create a stacked bar plot comparing the disease distributions of six regions.

```{r, eval=FALSE}
# Use geom_col() to create a stacked bar plot
# Use position = "fill" in geom_col()
# Assign correct x, y, and fill aesthetics

disease2016_summary %>% ggplot(aes(x=...,
                                   y=..., 
                                   fill=...) +
                        geom_...()
```


Let's clean up the plot a bit.

```{r, eval=FALSE}
# Create 'other' category for non-prevalent disease types.
# Assign the result to disease2016_summary2
# Four disease types should be "mumps", "pertussis", "measles", and "other"
disease2016_summary2  <- ...
```

```{r, eval=FALSE}
# Order disease types and regions for better visualisation
disease2016_summary2 <- disease2016_summary2 %>%
                        group_by(region, disease) %>%
                        summarise(cases = sum(cases), .groups = "drop") %>%
                        mutate(disease = factor(disease, 
                                                 levels = c("other","measles", 
                                                            "pertussis", "mumps")),
                                region = factor(region,
                                                levels = c("EUR", "SEAR",
                                                           "AFR", "AMR",
                                                           "WPR", "EMR")))
```


Create a publication quality stacked barplot

```{r, eval=FALSE}
# Assign correct x, y, and fill aesthetics
# Assign correct geom with appropriate position argument
# Use theme_economist()
disease2016_summary2 %>% ggplot(aes(x=..., y=..., fill=...)) +
  geom_... +
  theme_... +
  theme(legend.title = element_text(size=20, face="bold"),
        axis.title.x = element_blank(),
        axis.text.x = element_text(size=15, face="bold"),
        axis.text.y = element_text(face="bold.italic"),
        axis.title.y = element_text(size=15, face = "bold"),
        legend.text = element_text(size=15),
        legend.key.size = unit(1, 'cm')) +
  scale_fill_manual(values = rev(economist_pal()(4))) +
  labs(fill = "Disease", y = "Proportion of cases") +
  labs(title = "Disease distribution for six regions")
```

 Bar Plots are great to compare many categories at once.                
Also, Barplot should be used if the data follow some notion            
of stacking --- it simply means that you can add more stuff            
on top of the stack to increase it's height. Dollar value              
or Number of cases are great example of scenarios where               
such stacking principle holds true. Ratios or Percentiles              
should not be displayed using barplots. Any non-linear transformation 
 as well, e.g., we often take `log10()` or `sqrt()` transformations         
and they should not be displayed as bars.                              

### Q5. Gun Murder cases in 10 States
```{r, eval=FALSE}
# Get the top-10 states with most number of cases
murders_top10 <- ... %>%
                 slice_max(total, n=...)
```

Construct a bare bone barplot

```{r, eval=FALSE}
# Assign correct x and y aesthetics
murders_top10 %>% ggplot(aes(x = ..., y=...)) +
                  geom_col()
```

Now construc an enhanced barplot


```{r, eval=FALSE}
# Assign correct x and y aesthetics
# Use a colour of liking to fill the bars
# Flip the coordinates to make a horizontal bar plot --- use coord_flip()
# Use the minimal theme
murders_top10 %>% ggplot(aes(x = fct_reorder(..., ...),
                             y=...)) +
  geom_col(fill = ...) +
  coord_... +
  theme_... +
  theme(panel.grid.minor = element_blank(),
        axis.title = element_blank(),
        axis.text.y = element_text(face = "bold.italic", size=12),
        axis.text.x = element_text(face="bold", size = 12, 
                                   color="gray50"),
        plot.title = element_text(hjust=0.5, face="bold", color="red4", size=15),
        plot.subtitle =  element_text(hjust=0.5, face="bold.italic", color="red4", size=10)) +
  labs(title = "Top 10 US States in 2010", 
         subtitle = "According to gun related deaths")
```

  
Now let's look at some examples of **Point Chart**
First obtain the disease data for **WPR** region for the years **2006** and
**2016**.

```{r, eval=FALSE}
# Get the WPR data for the years 2006 and 2016 
disease_wpr <- disease_data %>%
               filter(...)

# You need to recreate country factor
disease_wpr$country <- factor(as.character(disease_wpr$...))

# Create country and year specific total cases
disease_wpr <- disease_wpr %>%
               group_by(..., ...) %>%
               summarise(cases = sum(...), .groups = "drop")

# Create two columns Cases2006 and Cases2016
# You can use pivot_wider() to achieve this
# Use year as name keys and cases as values
disease_wpr2 <- disease_wpr %>%
                pivot_wider(names_from = ...,
                            values_from = ...,
                            names_prefix = "Cases")
# Print disease_wpr2 to have a look at the data
...

# Filter out Countries with zero number of cases 
disease_wpr2 <- disease_wpr2 %>%
                filter(Cases2006 > 0, Cases2016 > 0)
```



### Q6. Plot log10(cases) for 2016 data.

```{r, eval=FALSE}
# Assign correct x, y aesthetic mappings
# Experiment with the coour of the points
# Create a horizontal barplot
# You can transform the y aesthetic to log10 scale using the scale_y_*()
# Use any theme you like --- I have put theme_bw() as a default

disease_wpr2 %>% ggplot(aes(x=fct_reorder(..., ...),
                            y=...)) +
                 geom_point(color= ...) +
                 coord_... +
                 scale_y_...(labels = comma) +
                 theme_bw() +
                 theme(axis.title.y = element_blank(),
                       panel.grid.major = element_line(color="white"),
                       panel.grid.minor = element_blank(),
                       axis.text.y = element_text(face = "bold.italic", size=10, color="blueviolet"),
                       axis.text.x = element_text(face = "bold", size=12, color="navy")) +
                 labs(y = "Number of disease cases (in log10 scale)",
                    title = "Ranking of African countries based on most to least number of cases in 2016")
```

### Q7. Create a point chart for log2-fold change using the WPR disease data for the years 2006 and 2016.


```{r, eval=FALSE}
# Create a column log2FoldChange using appropriate formula
disease_wpr3 <- disease_wpr2 %>% 
                mutate(log2FoldChange = log2(...))


# Create the Point Chart showing log2-fold-change values of African countries
# Assign appropriate x and y aesthetics
# Assign Correct geom_*
# Create a horizontal point chart --- use an appropriate coord_* function
# Experiment with themes --- as the default theme_bw() is given
# Add another geom_* layer to include a dashed-line as a visual anchor at 0 log2-fold-change value --- use geom_hline()


disease_wpr3 %>% ggplot(aes(x=...,
                            y=...)) +
  geom_... +
  coord_... +
  scale_y_continuous(labels = comma) +
  theme_bw() +
  geom_...(yintercept=0, linetype= ..., size=...) +
  theme(axis.title.y = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_text(face = "bold.italic", size=12),
        axis.text.x = element_text(face = "bold", size=12)) +
  labs(y = "Log2-fold change: log2(Case2016/Case2006)",
       title = "Ranking of African countries based on log2-fold changes")
```

Now let's look at some examples of distributional data. We are going to use the `iris` dataset for this last section.

```{r, eval=FALSE}
# Load iris data using data() function
data(...)

# Assign to a variable called iris
iris <- ...

# Inspect head of the data
head(...)

# Get the summary of the data
summary(...)

# Obtain the data only for the Setosa species 
setosa_df <- iris %>% filter(...)


# Create a Histogram using geom_histogram() of Sepal.Length
setosa_df %>%
  ggplot(aes(x=Sepal.Length)) +
  geom_...

# Create a Boxplot of sepal length data using geom_boxplot()
# Note geom_jitter() in the code creating the jittering of the points
setosa_df %>% ggplot(aes(x=Species, y=Sepal.Length, fill=Species)) +
  geom_... +
  scale_fill_viridis(discrete=TRUE, alpha=0.2) +
  geom_jitter(width=0.1, height=0, color="orange") +
  theme_bw()+
  theme(
    legend.position="none",
    plot.title = element_text(size=15),
    axis.title.y = element_text(size=15),
    axis.text.y = element_text(face="bold.italic", size=12),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.length.x = unit(0, 'cm'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    
  ) +
  ggtitle("Sepal length distribution of Setosa") +
  labs(y="Sepal length (cm)")
```


### Q8. Create Kernel density estimates of the Petal.Length densities for the three iris species.                                  


```{r, eval=FALSE}
# Use the correct x and fill aesthetics --- note we do not need y-aesthethics for density plot
# Use geom_density with alpha = 0.3
iris %>% ggplot(aes(x = ..., fill = ...)) +
         geom_...
```

### Q8. Create Kernel density estimates of the Petal.Length using three different bandwidth sizes: 0.58, (0.58/2), and (0.58/4)


```{r, eval=FALSE}
# Construct the density using default bandwidth of 0.58
iris %>% ggplot(aes(x = Petal.Length)) +
                geom_density(fill = "maroon") +
                ggtitle("Bandwidth = 0.58")

# Construct the density using bandwidth 0.58/2 --- use adjust = 0.5
iris %>% ggplot(aes(x = Petal.Length)) +
                       geom_density(fill = "maroon", adjust = ...) +
                       ggtitle("Bandwidth = (0.58)/2")

# Construct the density using bandwidth 0.58/4 --- use adjust = 0.25
iris %>% ggplot(aes(x = Petal.Length)) +
  geom_density(fill = "maroon", adjust = ...) +
  ggtitle("Bandwidth = (0.58)/4")
```


### Q9. Create Kernel density estimates of the Petal.Length for three species in the same plot.

```{r, eval=FALSE}
# Create species labels to use as annotations withing the plot
species_label <- data.frame(x=c(2.0, 4.5, 6.0), 
                            y=c(2.5, 1.1, 0.8), 
                            Species = c("Setosa", "Versicolor",
                                        "Virginica"))

# Create the Kernel Density Plot
iris %>% ggplot(aes(x = Petal.Length, fill = Species)) +
  geom_density(alpha = 0.3, color="black") +
  theme_bw() +
  scale_fill_viridis(discrete = TRUE) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  geom_text(data = species_label, aes(x=x, y=y, label=Species), 
            size = 4.5)
```


### Q10. Create Boxplots of Petal.Widths comparing three species.

```{r, eval=FALSE}
# Assign correct x, y, and fill aesthetics
# Use appropriate geom_* 
# Add jittered points
iris %>% ggplot(aes(x = ..., y=..., fill = ...)) +
   geom_...(color="white", outlier.size = 0.5) +
  geom_...(color="white", size=0.5, width=0.1, height=0) +
  scale_fill_viridis(discrete = TRUE) +
  theme_dark() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(face="bold", size=15),
        axis.title.y = element_text(face = "bold", size=15),
        axis.text.y = element_text(face="bold.italic", size=12,
                                   color = "gray50")) +
   ggtitle("Petal Width distributions of Iris species")
```


### Q11. Create **ridgeline** density plots for three iris species based on their Sepal.Length densities.

```{r, eval=FALSE}
# Assign correct x, y, and fill aesthetics.
# Use geom_density_ridges() to create the ridgeline plot

iris %>% ggplot(aes(y= ..., x=..., fill = ...)) +
  geom_... +
  scale_fill_viridis(discrete = TRUE) +
  theme_ridges() +
  labs(x = "Petal width (cm)") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_text(color = "gray50"),
        axis.text.y = element_text(size=15, face = "bold.italic", color="gray50"),
        legend.position = "none")
```
   


